\Large\textbf{General Stage Untangling with Applications to Algorithm Derivation}

### 2014-06-19 paper/semantics.tex
checkin at end of Kayvon's edits for the day.  Handing over to Nico.

\section{A Two-Stage Modal Language}

@@

In this section, we describe \lang, a simple two-stage language
used as the basis for defining and analyzing stage splitting.
%The grammar and type system for \lang\ is given in
%\ref{fig:grammar,fig:statics}.
Every expression in \lang\ has a statically known {\em stage},
either \bbone~or \bbtwo, which defines when it is fully reduced.
%\footnote{Typing judgments and context variables are annotated with stages.}
%
% in addition to a type.
% Just as a type describes \emph{which} values an expression reduces to,
%the expression's stage tells us \emph{when} it will be fully reduced. 
% FIXME why?
%The reason for a two-stage design (as opposed to \cite{davies96}, which allows an infinite number of stages) will be discussed later.
Stage-\bbone{} and stage-\bbtwo{} types include products, sums, functions, and base types.
In addition to these types, stage-\bbone{} expressions may also be of type $\fut A$,
representing an enclosed stage-\bbtwo{} expression of type $A$.
$\fut A$ terms must be treated opaquely by stage-\bbone{} code
as they cannot be reduced (to values with type $A$) until stage \bbtwo.

\needsfix{Although there is a strict one-way dependence between the two stages at the type
level}, stage-\bbone{} and stage-\bbtwo{} terms are defined mutually recursively:
the $\next$ construct embeds stage-\bbtwo{} expressions into stage \bbone,
whereas $\prev$ embeds stage-\bbone{} expressions into stage \bbtwo.
The $\next$ and $\prev$ constructs are the only expressions that interact with the stage of a term,
and we surround their arguments with braces in \lang\ syntax to clearly indicate stage boundaries within a program.

The grammar and type system for \lang\ is given in
\ref{fig:grammar,fig:statics}. (The figures annotate typing judgments and context variables with stages.)
Only $\fut$ and its introductory and eliminatory forms $\next$ and $\prev$ affect the stage
of a term or type.
%We formulated our typing judgments in the style of \cite{davies96}, where the
%whole judgment is annotated with a stage.  
%The grammar and type system for \lang\ is given in
%\ref{fig:grammar,fig:statics}.
% We annotate typing judgments and context variables with stages;
%This is made manifest as rules which are entirely abstract over stage.
%In addition to determining the stage, $\next$ and $\prev$ are the introduction and elimination forms for $\fut$ types.
Specifically, given an argument with type $A$ at stage \bbtwo, $\next$ forms a $\fut A$ at stage \bbone.  
%That is, it forms the promise of a future $A$ out of a construction for an $A$ at the next timestep.
Stage \bbtwo\ expressions can obtain the original stage \bbtwo\ argument via the $\prev$ construct.  
Since $\prev$ operates at stage \bbtwo, this ensures no violation of causality.
The $\pause$ construct serves to wrap stage \bbone\ integers for use in stage \bbtwo.  
Although it is possible to implement $\pause$ from other \lang\ features, 
we provide it as a core primitive to simplify our examples.

We note that our work does not consider the problem of {\em binding time analysis}---transforming an unstaged program into a staged \lang\ program by inserting appropriate $\next$, $\prev$, and $\pause$ constructs. We assume that binding time analysis is performed manually by a programmer or via automatic analysis of unstaged code to arrive at a valid \lang\ program used as input for subsequent splitting.

@@

Although the code for {\tt fexp} looks very much like the unstaged mathematical
definition above, it is in fact a staged program: the {\tt if} predicates and exponent
decomposition are all stage-\bbone\ terms, since they occur within $\prev$ blocks.
\needsfix{Essentially, the type system is sufficient to prove the observation that all of
the decomposition of the exponent can be done at the stage \bbone\ without
requiring disruption of the elegant functional structure of the code.}

A key attribute of the {\tt fexp} code example is the nesting of stage-\bbone\ and stage-\bbtwo\ expressions. Ordinary term evaluation always eliminates outermost redexes first, 
however in the case that stage \bbone\ expressions are contained inside
stage \bbtwo\ ones (such as in the example above), this strategy conflicts with the precept of staged execution: that all stage \bbone\ code be evaluated before the evaluation of stage \bbtwo\ code. Therefore, a suitably staged dynamic semantics for \lang\ must evaluate
\emph{all} the stage \bbone\ subexpressions of a term before its stage
\bbtwo\ subexpressions.

To understand how this behavior can be achieved, consider the following stage-\bbone\ term:
\begin{lstlisting} 
let x = (next {1+2}, 3+4) in
  next{ prev{#1 x} * prev{#1 x} * prev{hold(#2 x)} }
\end{lstlisting}
One option is to duplicate the contents of the first $\next$ expression, to produce
\begin{lstlisting}
(1+2) * (1+2) * 7
\end{lstlisting}

Here we are essentially treating $\next\{e\}$ as a stage \bbone{} value, so long
as all the $\prev$ subterms of $e$ are reduced. This is the approach taken by
\cite{davies96}.

@@

We think that this interpretation makes more sense from the point of view of cost, 
since the contents of a $\next$ always appear once in the residual, regardless of how many times the result is used.  
This is \emph{nearly} equivalent to the first expression, because they have different termination behavior 
when a non-terminating stage-\bbtwo~expression is declared but never used.

How do our semantics achieve this behavior? In the original term,
\begin{lstlisting}
let x = (next {1+2}, 3+4) in
  next{ prev{#1 x} * prev{#1 x} * prev{hold (#2 x)} }
\end{lstlisting}
we must start by evaluating the binding for $x$.  We can't evaluate inside the
$\next$ statement, but must still substitute some result into the body of the
\texttt{let}. We replace the contents of $\next$ with a variable and make an
explicit substitution, which then floats up to the top of the containing
$\prev$:
\begin{lstlisting} 
[1+2/x1]
let x = (next{x1}, 7) in
next{ prev{#1 x} * prev{#1 x} * prev{hold (#2 x)} }
\end{lstlisting}
We're now free to perform the stage-\bbone~substitution for $x$.  
\begin{lstlisting} 
[1+2/x1]
next{
	prev{#1 (next{x1}, 7)} *
	prev{#1 (next{x1}, 7)} *
	prev{hold (#2 (next{x1}, 7))}
}
\end{lstlisting}
To evaluate the remaining $\next$, we must first partially evaluate the body by finding all of the contained stage-\bbone~terms and reducing them:
\begin{lstlisting} 
[1+2/x1]
	next{ x1 * x1 * 7 }
\end{lstlisting}
The final step in evaluating the outer $\prev$ is reifying the contained substitutions into let statements, yielding
\begin{lstlisting} 
let x1 = 1+2 in
  x1 * x1 * 7
\end{lstlisting}

>>>>

\section{The \lang\ Language}

@@

In this section, we describe \lang, a simple two-stage language
used as the basis for defining and analyzing stage splitting.
%The grammar and type system for \lang\ is given in
%\ref{fig:grammar,fig:statics}.
Every expression in \lang\ has a statically known {\em stage},
either \bbone~or \bbtwo, which defines when it is fully reduced.
%\footnote{Typing judgments and context variables are annotated with stages.}
%
% in addition to a type.
% Just as a type describes \emph{which} values an expression reduces to,
%the expression's stage tells us \emph{when} it will be fully reduced. 
% FIXME why?
%The reason for a two-stage design (as opposed to \cite{davies96}, which allows an infinite number of stages) will be discussed later.
Stage-\bbone\ and stage-\bbtwo\ types include products, sums, functions, and base types.
In addition to these types, stage-\bbone\ expressions may also be of type $\fut A$,
representing an enclosed stage-\bbtwo\ expression of type $A$.
(The type system of \lang\ is adopted directly from that of the modal language \lamCircle\,\cite{davies96}, but is restricted to two stages for simplicity.) $\fut A$ terms must be treated opaquely by stage-\bbone\ code
as they cannot be reduced (to values with type $A$) until stage \bbtwo. Like in \lamCircle\, there is no reciprocal modality.

Although there is a strict one-way dependence between the two stages at the type
level, stage-\bbone\ and stage-\bbtwo\ terms are defined mutually recursively:
the $\next$ construct embeds stage-\bbtwo\ expressions into stage \bbone,
whereas $\prev$ embeds stage-\bbone\ expressions into stage \bbtwo.
The $\next$ and $\prev$ constructs are the only expressions that interact with the stage of a term,
and we surround their arguments with braces in \lang\ syntax to clearly indicate stage boundaries within a program.

The grammar and type system for \lang\ is given in
\ref{fig:grammar,fig:statics}. (Typing judgments and context variables are annotated with stages.)
Only $\fut$ and its introductory and eliminatory forms $\next$ and $\prev$ affect the stage
of a term or type.
%We formulated our typing judgments in the style of \cite{davies96}, where the
%whole judgment is annotated with a stage.  
%The grammar and type system for \lang\ is given in
%\ref{fig:grammar,fig:statics}.
% We annotate typing judgments and context variables with stages;
%This is made manifest as rules which are entirely abstract over stage.
%In addition to determining the stage, $\next$ and $\prev$ are the introduction and elimination forms for $\fut$ types.
Specifically, given an argument with type $A$ at stage \bbtwo, $\next$ forms a $\fut A$ at stage \bbone.  
%That is, it forms the promise of a future $A$ out of a construction for an $A$ at the next timestep.
Stage \bbtwo\ expressions can obtain the original stage \bbtwo\ argument via the $\prev$ construct.  
Since $\prev$ operates at stage \bbtwo, this ensures no violation of causality\,\cite{cave14}.
The $\pause$ construct serves to wrap stage \bbone\ integers for use in stage \bbtwo.  
Although it is possible to implement $\pause$ from other \lang\ features, 
we provide it as a core primitive to simplify our examples.

@@

Although the code for {\tt fexp} looks very much like the unstaged mathematical
definition above, it is in fact a staged program: the {\tt if} predicates and exponent
decomposition are all stage-\bbone\ terms, since they occur within $\prev$ blocks.
%\needsfix{Essentially, the type system is sufficient to prove the observation that all of
%the decomposition of the exponent can be done at the stage \bbone\ without
%requiring disruption of the elegant functional structure of the code.}
We note that our work does not consider the problem of {\em binding time analysis}---transforming an unstaged program into a staged \lang\ program by inserting appropriate $\next$, $\prev$, and $\pause$ constructs. We assume that binding time analysis is performed manually by a programmer (as done above) or via automatic analysis of unstaged code to arrive at a valid \lang\ program used as input for subsequent splitting.

A key attribute of the {\tt fexp} code example is the nesting of stage-\bbone\ and stage-\bbtwo\ expressions. Ordinary term evaluation eliminates outermost redexes first, 
however in the case that stage \bbone\ expressions are contained inside
stage \bbtwo\ ones (such as in the example above), this strategy conflicts with the precept of staged execution: that all stage \bbone\ code be evaluated before the evaluation of stage \bbtwo\ code. Therefore, a suitably staged dynamic semantics for \lang\ must evaluate
\emph{all} the stage \bbone\ subexpressions of a term before its stage
\bbtwo\ subexpressions.

To understand how this behavior can be achieved, consider evaluation of the stage-\bbone\ term below. (The term contains nested $\next$ and $\prev$ blocks.):
\begin{lstlisting} 
let x = (next {1+2}, 3+4) in
  next{ prev{#1 x} * prev{#1 x} * prev{hold(#2 x)} }
\end{lstlisting}
One option is to duplicate the contents of the first $\next$ expression, to produce
\begin{lstlisting}
(1+2) * (1+2) * 7
\end{lstlisting}

Here we are essentially treating $\next\{e\}$ as a stage-\bbone\ value, so long
as all the $\prev$ subterms of $e$ are reduced. This is the approach taken by
\cite{davies96}.

@@

% We think that this interpretation makes more sense from the point of view of cost, 
% since the contents of a $\next$ always appear once in the residual, regardless of how many times the result is used.  
% This is \emph{nearly} equivalent to the first expression, because they have different termination behavior 
% when a non-terminating stage-\bbtwo~expression is declared but never used.

To achieve this behavior, we must start by evaluating the binding for $x$.  We can't evaluate inside the
$\next$ statement, but must still substitute some result into the body of the
\texttt{let}. We replace the contents of $\next$ with a variable and make an
explicit substitution, which then floats up to the top of the containing
$\prev$:
\begin{lstlisting} 
[1+2/x1]
let x = (next{x1}, 7) in
next{ prev{#1 x} * prev{#1 x} * prev{hold (#2 x)} }
\end{lstlisting}
We're now free to perform the stage-\bbone~substitution for $x$.  
\begin{lstlisting} 
[1+2/x1]
next{
	prev{#1 (next{x1}, 7)} *
	prev{#1 (next{x1}, 7)} *
	prev{hold (#2 (next{x1}, 7))}
}
\end{lstlisting}
To evaluate the remaining $\next$, we must first partially evaluate the body by finding all of the contained stage-\bbone~terms and reducing them:
\begin{lstlisting} 
[1+2/x1]
	next{ x1 * x1 * 7 }
\end{lstlisting}
The final step in evaluating the outer $\prev$ is reifying the contained substitutions into let statements, yielding
\begin{lstlisting} 
let x1 = 1+2 in
  x1 * x1 * 7
\end{lstlisting}
